---
title: "minPanel"
author: Tingting Zhan
date: "`r Sys.Date()`"
format: 
  html:
    page-layout: full
    html-math-method: katex
number-sections: true
toc: true
toc-location: left
toc-depth: 4
toc-title: ''
editor: source
knitr:
  opts_chunk: 
    collapse: true
    comment: "#" 
bibliography: minPanel.bib
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette of package **`minPanel`** ([Github](https://github.com/tingtingzhan/minPanel)) presents ...

<!--
To cite this methodology, please use @Zezulinski25.
-->


## Prerequisite

Package **`minPanel`** requires R version 4.5.0 (released 2025-04-11) or higher ([macOS](https://cran.r-project.org/bin/macosx/), [Windows](https://cran.r-project.org/bin/windows/base/), [Linux](https://cran.r-project.org/bin/linux/)).

```{r}
#| code-fold: true
#| code-summary: "Environment on author's computer"
#| label: author-env
Sys.info()[c('sysname', 'release', 'machine')]
R.version
```


Experimental (and maybe unstable) features are released *extremely frequently* to [Github](https://github.com/tingtingzhan/minPanel). <!--Stable releases to [CRAN](https://CRAN.R-project.org/package=minPanel) are typically updated every 2 to 3 months.-->

```{r}
#| eval: false
remotes::install_github('tingtingzhan/minPanel')
```


```{r}
#| eval: false
#| echo: false
utils::install.packages('minPanel')
```

## Getting Started

Examples in this vignette require that the `search` path has

```{r}
library(minPanel)
library(ggplot2)
library(flextable)
```



# Data Set

Study data should be described in a peer-reviewed manuscript in the following language.

> `<Biology>` samples from $n_1$ `<positive.subjects>` were obtained from our collaborators `<institutionA-1>`, `<institutionA-2>` under approved IRB protocols.  `<Biology>` samples from $n_0$ `<negative.subjects>` from `<institutionB-1>`, `<institutionB-2>` were used as controls.  Demographic information of human subjects is shown in `<table>` and `<supplementary materials>`.

Data from @Zezulinski25 is included in package **`minPanel`**.

@lst-m1 shows that the data object *`m1`* is a `logical` `matrix` of `r nrow(m1)` variants in `r ncol(m1)` early & late stage, i.e., *positive*, patients.

```{r}
#| lst-cap: 'Number of variants and positive patients'
#| lst-label: lst-m1
m1 |> 
  dim()
```

```{r}
#| lst-cap: 'A snippet of `m1` (@lst-m1)'
#| lst-label: lst-m1_head
m1 |> 
  head(n = c(3L, 3L))
```

@lst-m0 shows that the data object *`m0`* is a `logical` `matrix` of the same `r nrow(m0)` variants in `r ncol(m0)` UPC, i.e., *negative*, patients.

```{r}
#| lst-cap: 'Number of variants and negative patients'
#| lst-label: lst-m0
m0 |> 
  dim()
```

```{r}
#| lst-cap: 'A snippet of `m0` (@lst-m0)'
#| lst-label: lst-m0_head
m0 |> 
  head(n = c(3L, 3L))
```

# Panel of Variant-Signatures {#sec-panel}

@lst-panel consolidates the variants that identify the same set of positive subjects in @lst-m1 and the same set of negative subjects in @lst-m0.  In other words, these variants are considered identical in terms of @lst-m1 and @lst-m0.  Therefore they are consolidated into one *variant-signature*.

```{r}
#| lst-label: lst-panel
#| lst-cap: 'Variant signatures in @lst-m1 and @lst-m0'
(m = panel(m1 = m1, m0 = m0))
```

@lst-panel consolidates the `r nrow(m1)` variants in @lst-m1 and @lst-m0 into `r length(m@id)` variant-signatures.  For example, the variants below (@lst-panel_eg) identify the same set of positive and negative subjects, thus they are consolidated as one signature.

```{r}
#| code-fold: true
#| lst-label: lst-panel_eg
#| lst-cap: 'e.g., variants consolidated in one variant-signature'
m@id[[8L]] |>
  cat(sep = '\n')
```


# Ordered Panel {#sec-select-panel}

Methodology should be described in a peer-reviewed manuscript in the following language.

> An *ordered panel* of variant-signatures is determined using the following steps: (1) exclude the signatures that identify none of the `<positive.subjects>` *per signature*; (2) exclude the signatures that identify `1`-or-more `<negative.subjects>` *per signature* (e.g., @lst-p0); (3) sort the remaining signatures by the number of `<positive.subjects>` identified *per signature*, from most to least; (4) excludethe signatures that identify no more `<positive.subjects>` than all the preceding signatures.  Such panel of signatures is ***ordered***, i.e., a researcher may reduce the size of the ordered-panel by keeping the signatures in the top of the ordered-panel and dropping those in the bottom of the ordered-panel.  `<Figure>` (e.g., @fig-consort0) presents the [Consolidated Standards of Reporting Trials (CONSORT)](https://en.wikipedia.org/wiki/Consolidated_Standards_of_Reporting_Trials) diagram of the creation of an ordered panel.  `<Table>` (e.g., @lst-flextable_panel) summarizes the true and false positive rates per signature, as well as per ordered sub-panel selected by a researcher. The ordered-panel methodology is implemented in R (@R) package **`minPanel`** (citation coming soon).

## `'False(+) ≤0'`



```{r}
#| lst-label: lst-x0_create
#| lst-cap: 'Ordered panel *`p0`*: controlled at `0`-or-less `False(+)` per signature'
p0 = m |>
  subset(subset = sum1 > 0L) |>
  subset(subset = sum0 <= 0L, append.label = TRUE) |>
  sort_by(y = ~ sum1, decreasing = TRUE) |>
  subset(subset = diff(cumsum1) > 0L) |>
  append_label(info = 'cumsum0')
```


The pipeline @lst-x0_create consists of five steps.

1. The first call to `subset()` (Appendix @sec-subset) selects the signatures that identify at least one positive subjects *per signature* (Appendix @sec-positive).
2. The second call to `subset()` selects the signatures that identify none of the negative subjects *per signature* (Appendix @sec-positive).  The [*per-signature* criterion]{style="background-color: #FFFF00"} *`'False(+) ≤0/31'`* is `append`ed to the `@label` of the output.
3. The call to `sort_by()` (Appendix @sec-sortby) sorts the signatures by their number of positive subjects identified *per signature* (Appendix @sec-positive), in `decreasing` order.
4. The last call to `subset()` identifies the signatures, so that each one signature identifies at least one more positive subjects than all the preceding signatures (Appendix @sec-positive).
5. The call to `append_label()` (Appendix @sec-appendlabel) summarizes the number of negative subjects identified *as a panel of signatures* (Appendix @sec-positive).  The [*panel-wise* criterion]{style="background-color: #FFFF00"} *`'panelFalse(+) ≤0/31'`* is `append`ed to the `@label` of the output.

@lst-p0 prints the vital information of the ordered panel @lst-x0_create.

```{r}
#| lst-label: lst-p0
#| lst-cap: 'Ordered panel *`p0`*: controlled at `0`-or-less `False(+)` per signature'
p0
```



@fig-consort0 (@lst-consort_panel) presents a Consolidated Standards of Reporting Trials (CONSORT) diagram of the creation of the ordered panel @lst-p0.



```{r}
#| lst-label: lst-consort_panel
#| lst-cap: 'CONSORT Diagram of ordered panel @lst-p0'
#| fig-height: 2.5
#| fig-width: 5
#| fig-align: left
#| fig-cap: 'CONSORT Diagram of Ordered Panel @lst-p0'
#| label: fig-consort0
p0 |>
  consort.panel()
```

@lst-flextable_panel displays the variant-signatures in the ordered panel @lst-p0, and summarizes the true and false positive rates per signature, as well as per ordered sub-panel selected by a researcher.  Those variant-signatures that contain more than one variants are highlighted in [light-yellow]{style="background-color: lightyellow"} color.

```{r}
#| lst-label: lst-flextable_panel
#| lst-cap: 'Variant-signatures in ordered panel @lst-p0'
#| tbl-cap: 'Variant-Signatures in Ordered Panel'
p0 |>
  as_flextable()
```


Similarly, we relax the constraint on `False(+)`-per-signature and create @lst-p1 - @lst-p3.

## `'False(+) ≤1'`

```{r}
#| lst-label: lst-p1
#| lst-cap: 'Ordered panel *`p1`*: controlled at `1`-or-less `False(+)` per signature'
p1 = m |>
  subset(subset = sum1 > 0L) |>
  subset(subset = sum0 <= 1L, append.label = TRUE) |>
  sort_by(y = ~ sum1, decreasing = TRUE) |>
  subset(subset = diff(cumsum1) > 0L) |>
  append_label(info = 'cumsum0')
p1
```

```{r}
p1 |>
  as_flextable()
```


## `'False(+) ≤2'`

```{r}
#| lst-label: lst-p2
#| lst-cap: 'Ordered panel *`p2`*: controlled at `2`-or-less `False(+)` per signature'
p2 = m |>
  subset(subset = sum1 > 0L) |>
  subset(subset = sum0 <= 2L, append.label = TRUE) |>
  sort_by(y = ~ sum1, decreasing = TRUE) |>
  subset(subset = diff(cumsum1) > 0L) |>
  append_label(info = 'cumsum0')
p2
```

```{r}
p2 |>
  as_flextable()
```


## `'False(+) ≤3'`

```{r}
#| lst-label: lst-p3
#| lst-cap: 'Ordered panel *`p3`*: controlled at `3`-or-less `False(+)` per signature'
p3 = m |>
  subset(subset = sum1 > 0L) |>
  subset(subset = sum0 <= 3L, append.label = TRUE) |>
  sort_by(y = ~ sum1, decreasing = TRUE) |>
  subset(subset = diff(cumsum1) > 0L) |>
  append_label(info = 'cumsum0')
p3
```

```{r}
p3 |>
  as_flextable()
```


# Ordered Panel List

@lst-panellist collects multiple ordered panels.

```{r}
#| lst-label: lst-panellist
#| lst-cap: 'Ordered panel list containing @lst-p0 - @lst-p3'
(z = panellist(p0, p1, p2, p3))
```



## Operating Characteristics

The operating characteristics should be described in a peer-reviewed manuscript in the following language.

> `<Figure_oc>` presents the operating characteristic curves of multiple ordered panels of variant-signatures.  The $x$-axis is the number of variant-signatures that are selected from each ordered panel by the investigator.  The $y$-axis is the percentage of `<positive.subjects>` identified, i.e., the sensitivity, for each selected number of variant-signatures.  An operating characteristics curve moves towards the upper-left corner to indicate a better performance, i.e., higer sensitivity with lower number of variant-signatures selected from the ordered panel. 
>
> `<Figure_roc>` presents the *pseudo* receiver operating characteristic (ROC) curves of multiple ordered panels of variant-signatures. The $x$-axis is the non-specificity for each selected number of variant-signatures.  The $y$-axis is the percentage of `<positive.subjects>` identified, i.e., the sensitivity, for each selected number of variant-signatures. 
This is a *pseudo* ROC curve, as each sensitivity-non-specificity pair on this curve corresponds to a different selected number of variant-signatures.

```{r}
#| echo: false
#| eval: false
scales::pal_hue()(n = 4L)
```

@fig-oc_panellist (@lst-oc_panellist) and @fig-roc_panellist (@lst-roc_panellist) visualize the operating characteristic and the  *pseudo* receiver operating characteristic curves of the ordered panel-list @lst-panellist.  It is concluded that the [**ordered panel @lst-p1**]{style="color: #7CAE00"}, controlled at a per-signature false positive of `<=1/31`, is the optimal among the four ordered panels.  

Compared to the [**ordered panel @lst-p1**]{style="color: #7CAE00"},

-   the [ordered panel @lst-p0]{style="color: #F8766D"}, controlled at a more stringent criterion of per-signature false positive of `<=0/31`, has 
    -   lower sensitivity, assuming the same number of variant-signatures selected (@fig-oc_panellist);  
    -   lower panel-wise non-specificity (@fig-roc_panellist);
-   the [ordered panel @lst-p2]{style="color: #00BFC4"} and [ordered panel @lst-p3]{style="color: #C77CFF"}, controlled at less stringent criteria of per-signature false positive of `<=2/31` and `<=3/31`, respectively, have 
    -   almost identical sensitivity, assuming the same number of variant-signatures selected (@fig-oc_panellist);
    -   higher panel-wise non-specificity (@fig-roc_panellist).

```{r}
#| lst-label: lst-oc_panellist
#| lst-cap: 'Operating characteristics of ordered panel list @lst-panellist'
#| fig-height: 3
#| fig-width: 6
#| fig-align: left
#| fig-cap: 'Operating Characteristics of Ordered Panel List @lst-panellist'
#| label: fig-oc_panellist
z |> 
  autoplot()
```

```{r}
#| lst-label: lst-roc_panellist
#| lst-cap: 'Pseudo receiver operating characteristics of ordered panel list @lst-panellist'
#| fig-height: 3
#| fig-width: 6
#| fig-align: left
#| fig-cap: 'Pseudo Receiver Operating Characteristics of Ordered Panel List @lst-panellist'
#| label: fig-roc_panellist
z |> 
  autoplot(which = 'roc')
```

<!--
## Selected Signatures

@lst-flextable_panellist displays the variant-signatures selected in each ordered panel of the panel-list @lst-panellist.  The variant-signatures that contain more than one variants are highlighted using the [light-yellow]{style="background-color: lightyellow"} color.

```{r}
#| lst-label: lst-flextable_panellist
#| lst-cap: 'Variant-signatures in each ordered panel of the panel-list @lst-panellist'
z |> 
  as_flextable()
```

-->

# Appendix

## About {#sec-about}

R terminology may differ from that commonly used in mathematics and statistics. For explanations and references to the terms and abbreviations used in this vignette, please refer to @sec-terms.

### Environment {#sec-envir}

This vignette is created under `r R.version$version.string` with the aid of packages **`knitr`** [@knitr, v`r packageVersion('knitr')`], **`quarto`** [@quarto, v`r packageVersion('quarto')` with [Quarto](https://quarto.org/docs/get-started/) v`r quarto::quarto_version()`] and **`rmarkdown`** [@rmarkdown, v`r packageVersion('rmarkdown')`]. Documentation is supported by package **`roxygen2`** [@roxygen2, v`r packageVersion('roxygen2')`].

An Integrated Development Environment (IDE), e.g., [RStudio](https://posit.co/download/rstudio-desktop/) [@RStudio] or [Positron](https://positron.posit.co/download.html), is not required, but highly recommended.

### Dependency {#sec-depend}

Package **`minPanel`** `Imports` packages

-   **`cli`** [@cli, v`r packageVersion('cli')`], for attractive command line interfaces
-   **`consort`** [@consort, v`r packageVersion('consort')`], for Consolidated Standards of Reporting Trials (CONSORT) diagram
-   **`flextable`** [@flextable, v`r packageVersion('flextable')`], for flexible table
-   **`ggplot2`** [@ggplot2], **`patchwork`** [@patchwork], for visualization
-   `r '\U1f5dd'` **`matrixStats`** [@matrixStats, v`r packageVersion('matrixStats')`], key dependency, for matrix algebra
-   `r '\U1f5dd'` **`methods`** shipped with vanilla `r R.version.string`, key dependencies, for `S4` generic function and methodes
-   **`scales`** [@scales, v`r packageVersion('scales')`], for colour pallet


## `panel`

The function `panel()` (@sec-panel) returns an object of `S4` class `'panel'`, with `slot`s

-   `@m1`, a `logical` `matrix`, signatures on rows and positive subjects on columns
-   `@m0`, a `logical` `matrix`, signatures on rows and negative subjects on columns
-   `@id`, a `list` of `character` `vector`s.  Each `character` `vector` contains the variants in a signature. 
-   `@label`, an optional `character` scalar

The `S4` method of `show` to `signature = 'panel'` prints the vital information, e.g., the number of signatures, and the number of positive and negative subjects.

### (Cumulative) True & False Positives {#sec-positive}

The function `sum1()` finds the number of true positives per signature, i.e., the `rowSums` of the slot `@m1`.

The function `sum0()` finds the number of false positives per signature, i.e., the `rowSums` of the slot `@m0`.

Functions `sum1()` and `sum0()` return 
an `integer` `vector`.
 
```{r}
# example
```


The function `cumsum1()` finds the number of true positives, by a sub-panel of the first $i$ signatures.

The function `cumsum0()` finds the number of false positives, by a sub-panel of the first $i$ signatures.

The function `cumsum1()` and `cumsum0()` return
a not-strictly increasing `integer` `vector`.

```{r}
# example
```


### Subset {#sec-subset}  

The `S3` method `subset.panel()` 🚧

#### by Panel-Wise False Positives

Panel *`p1`* in @sec-select-panel, controlled at `sum0 <= 1L` per signature, has a substantial panel-wise false positive rate

```{r}
p1
```

User may impose an additional criterion of panel-wise false positive of 5 or less out of 31 negative subjects.

```{r}
#| code-fold: true
#| code-summary: '**Example**: additional criteria of panel-wise false positive `<=5`'
x1a = m |>
  subset(subset = sum1 > 0L) |>
  subset(subset = sum0 <= 1L, append.label = TRUE) |>
  sort_by(y = ~ sum1, decreasing = TRUE) |>
  subset(subset = diff(cumsum1) > 0L) |>
  subset(subset = cumsum0 <= 5L, append.label = TRUE)
```

```{r}
x1a
```

@fig-panel-a shows ..

```{r}
#| code-fold: true
#| code-summary: '**Figure**: operating characteristics of *`p0`*, *`p1`* and *`x1a`*'
#| fig-height: 3
#| fig-width: 6
#| fig-align: left
#| label: fig-panel-a
#| fig-cap: 'Operating characteristics of *`p0`*, *`p1`* and *`x1a`*'
panellist(p0, p1, x1a) |>
  autoplot()
```


### Sort-By {#sec-sortby}

The `S3` method `sort_by.panel()` 🚧

### Append Label {#sec-appendlabel}

The function `append_label()` 🚧

### CONSORT Diagram {#sec-consort_panel}


The function `consort.panel()` creates a CONSORT diagram of the creation of a `panel`, using package **`consort`** [@consort].  

The function `consort.panel()` is a wrapper of the function `consort::consort_plot()`, which returns an object of class `'consort'`.

Examples using function `consort.panel()` are provided in @sec-select-panel.

## `panellist`

### CONSORT Diagram {#sec-consort_panellist}

The function `consort.panellist()` organizes the CONSORT diagrams of each `panel` member in a `panellist` in a collage, using packages **`consort`** [@consort] and **`patchwork`** [@patchwork].  

@fig-consort presents side-by-side the selection of signatures in *`p0`*, *`p1`*, and *`p2`*.

```{r}
#| code-fold: true
#| code-summary: '**Figure**: CONSORT diagrams of *`p0`*, *`p1`*, and *`p2`*'
#| fig-height: 3.5
#| fig-width: 10
#| fig-align: left
#| fig-cap: 'CONSORT Diagram of `panellist`'
#| label: fig-consort
z[1:3] |> 
  consort.panellist(nrow = 1L)
```


# What We Don't Do

## `subset.panel()`

### by Multiple Criteria

too difficult to parse..

```{r}
#| eval: false
m |> 
  subset.panel(subset = (sum1 > 0L) & (sum0 <= 3L))
```


# Terms & Abbreviations {#sec-terms}


@tbl-Rterm presents a comprehensive glossary of R terms and abbreviations used in this vignette, which is folded to avoid overwhelming the reader with its excessive length.

R terminology and nomenclature could be *drastically different* from that of mathematics and statistics. Readers are *strongly advised* to read closely from the reference links in @tbl-Rterm, most of which point to webpages on [search.r-project.org](https://search.r-project.org), [cran.r-project.org](https://cran.r-project.org), [adv-r.hadley.nz](https://adv-r.hadley.nz) and [en.wikipedia.org](https://en.wikipedia.org).

<details>

<summary>@tbl-Rterm: Terms & Abbreviations in R</summary>

| Term / Abbreviation | Description |
|:-----------------------------------|:-----------------------------------|
| [moved] | [moved] |

: Terms & Abbreviations in R {#tbl-Rterm}

</details>

{{< pagebreak >}}



# References {.appendix}

::: {#refs}
:::
