---
title: "minPanel"
author: Tingting Zhan
date: "`r Sys.Date()`"
format: 
  html:
    page-layout: full
    html-math-method: katex
number-sections: true
toc: true
toc-location: left
toc-depth: 4
toc-title: ''
editor: source
knitr:
  opts_chunk: 
    collapse: true
    comment: "#" 
bibliography: minPanel.bib
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette of package **`minPanel`** ([Github](https://github.com/tingtingzhan/minPanel)) presents ...

<!--
To cite this methodology, please use @Zezulinski25.
-->


## Prerequisite

Package **`groupedHyperframe`** requires R version 4.5.0 (released 2025-04-11) or higher ([macOS](https://cran.r-project.org/bin/macosx/), [Windows](https://cran.r-project.org/bin/windows/base/), [Linux](https://cran.r-project.org/bin/linux/)).

```{r}
#| code-fold: true
#| code-summary: "Environment on author's computer"
#| label: author-env
Sys.info()[c('sysname', 'release', 'machine')]
R.version
```


Experimental (and maybe unstable) features are released *extremely frequently* to [Github](https://github.com/tingtingzhan/minPanel). <!--Stable releases to [CRAN](https://CRAN.R-project.org/package=minPanel) are typically updated every 2 to 3 months.-->

```{r}
#| eval: false
remotes::install_github('tingtingzhan/minPanel')
```


```{r}
#| eval: false
#| echo: false
utils::install.packages('minPanel')
```

## Getting Started

Examples in this vignette require that the `search` path has

```{r}
library(minPanel)
library(ggplot2)
library(flextable)
```



# Data Set

Study data should be described in a peer-reviewed manuscript in the following language.

> `<Biology>` samples from $n_1$ `<positive.subjects>` were obtained from our collaborators `<institutionA-1>`, `<institutionA-2>` under approved IRB protocols.  `<Biology>` samples from $n_0$ `<negative.subjects>` from `<institutionB-1>`, `<institutionB-2>` were used as controls.  Demographic information of human subjects is shown in `<table>` and `<supplementary materials>`.

Data from @Zezulinski25 is included in package **`minPanel`**.

Data object *`m1`* is a `logical` `matrix` of `r nrow(m1)` variants in `r ncol(m1)` early & late stage, i.e., *positive*, patients.

```{r}
m1 |> 
  dim()
```

```{r}
m1 |> 
  head(n = c(3L, 3L))
```

Data object *`m0`* is a `logical` `matrix` of the same `r nrow(m0)` variants in `r ncol(m0)` UPC, i.e., *negative*, patients.

```{r}
m0 |> 
  dim()
```

```{r}
m0 |> 
  head(n = c(3L, 3L))
```

# Panel of Variant-Collections {#sec-panel}

Function `panel()` identifies the variants that identifies the same set of positive subjects in *`m1`*, and the same set of negative subjects in *`m0`*.  These variants are considered identical in terms of *`m1`* and *`m0`*.  Therefore they are put in one **variant-collection**.  We use *"collection"* for short in this vignette.

```{r}
m = panel(m1 = m1, m0 = m0)
```



```{r}
m
```

In other words, the `r nrow(m1)` variants in both *`m1`* and *`m0`* are consolidated into `r length(m@id)` collections.  For example, the variants below identify the same set of positive and negative subjects, thus they are consolidated as one collection.

```{r}
#| code-fold: true
#| code-summary: e.g., variants consolidated in one collection
m@id[[8L]] |>
  cat(sep = '\n')
```


# Selected Ordered Panel {#sec-select-panel}

Methodology should be described in a peer-reviewed manuscript in the following language.

> A minimal panel of variants-collections is determined using the following steps: (1) we exclude the collections that identify none of the `<positive.subjects>` *per collection*; (2) we exclude the collections that identify at least `1` of the `<negative.subjects>` *per collection*; (3) we sort the remaining collections by the number of `<positive.subjects>` identified *per collection*, from most to least; (4) we exclude  the collections that identify no more `<positive.subjects>` than all the preceding collections.  Such panel of collections is ***ordered***, i.e., a researcher may reduce the size of the panel by keeping the collections in the top of the panel and dropping those in the bottom of the panel.  `<Figure>` is the Consolidated Standards of Reporting Trials (CONSORT) diagram of this selection procedure. This methodology is implemented in R (@R) package **`minPanel`** (citation coming soon).


```{r}
x0 = m |>
  subset(subset = true_positive > 0L) |>
  subset(subset = false_positive <= 0L, append.label = TRUE) |>
  sort_by(y = ~ true_positive, decreasing = TRUE) |>
  subset(subset = diff(cum_true_positive) > 0L) |>
  append_label(info = 'cum_false_positive')
```


The pipeline above consists of five steps.

1. The first call to `subset()` (Appendix @sec-subset) selects the collections that identify at least one positive subjects *per collection* (Appendix @sec-positive).
2. The second call to `subset()` selects the collections that identify none of the negative subjects *per collection* (Appendix @sec-positive).  The [*per-collection* criterion]{style="background-color: #FFFF00"} *`'False(+) ≤0/31'`* is `append`ed to the `@label` of the output.
3. The call to `sort_by()` (Appendix @sec-sortby) sorts the collections by their number of positive subjects identified *per collection* (Appendix @sec-positive), in `decreasing` order.
4. The last call to `subset()` identifies the collections, so that each one collection identifies at least one more positive subjects than all the preceding collections (Appendix @sec-positive).
5. The call to `append_label()` (Appendix @sec-appendlabel) summarizes the number of negative subjects identified *as a panel of collections* (Appendix @sec-positive).  The [*panel-wise* criterion]{style="background-color: #FFFF00"} *`'panelFalse(+) ≤0/31'`* is `append`ed to the `@label` of the output.


```{r}
x0
```

Function `consort.panel()` (Appendix @sec-consort_panel) presents a Consolidated Standards of Reporting Trials (CONSORT) diagram of the creation of a selected panel.  @fig-consort0 shows the creation of *`x0`*.

```{r}
#| fig-height: 2.5
#| fig-width: 5
#| fig-align: left
#| fig-cap: 'CONSORT Diagram of *`x0`*'
#| label: fig-consort0
x0 |>
  consort.panel()
```


Similarly, we relax the constraint on `false_positive`-per-collection and create

```{r}
#| code-fold: true
#| code-summary: 'A `panel` *`x1`*: controlled at `false_positive <= 1L` per collection'
x1 = m |>
  subset(subset = true_positive > 0L) |>
  subset(subset = false_positive <= 1L, append.label = TRUE) |>
  sort_by(y = ~ true_positive, decreasing = TRUE) |>
  subset(subset = diff(cum_true_positive) > 0L) |>
  append_label(info = 'cum_false_positive')
```

```{r}
x1
```


```{r}
#| code-fold: true
#| code-summary: 'A `panel` *`x2`*: controlled at `false_positive <= 2L` per collection'
x2 = m |>
  subset(subset = true_positive > 0L) |>
  subset(subset = false_positive <= 2L, append.label = TRUE) |>
  sort_by(y = ~ true_positive, decreasing = TRUE) |>
  subset(subset = diff(cum_true_positive) > 0L) |>
  append_label(info = 'cum_false_positive')
x2
```


```{r}
x2
```


```{r}
#| code-fold: true
#| code-summary: 'A `panel` *`x3`*: controlled at `false_positive <= 3L` per collection'
x3 = m |>
  subset(subset = true_positive > 0L) |>
  subset(subset = false_positive <= 3L, append.label = TRUE) |>
  sort_by(y = ~ true_positive, decreasing = TRUE) |>
  subset(subset = diff(cum_true_positive) > 0L) |>
  append_label(info = 'cum_false_positive')
x3
```

```{r}
x3
```



# Operating Characteristics

Function `panellist()` collects multiple `panel`s.

```{r}
z = panellist(x0, x1, x2, x3)
```

The `S3` method dispatch `base::print.listof()` displays a `panellist`.

```{r}
#| code-fold: true
#| code-summary: 'A `panellist` *`z`*'
z
```


## Visualization

The operating characteristics should be described in a peer-reviewed manuscript in the following language.

> `<Figure>` presents the operating characteristics of multiple ordered-panels of variants-collections.  The $x$-axis is the number of collections from the top of each ordered-panel.  The $y$-axis is the percentage of `<positive.subjects>` identified.  An operating characteristics curve moves towards the upper-left corner to indicate a better performance, i.e., higer percentage of `<positive.subjects>` identified with lower number of collections in the ordered-panel. Investigators should evaluate multiple operating characteristics curves by their control on number of `<false.subjects>` identified.


The `S3` method dispatches `autoplot.panellist()` and `autolayer.panellist()` visualize the operating characteristics of panels *`x0`*, *`x1`*, *`x2`* and *`x3`* in @fig-panellist.  

@fig-panellist concludes that the panel *`x1`*, controlled at per-collection false positive of `<=1/31` (colored [**green**]{style="color: #7CAE00"}), is the optimal among the four panels, as

-   Panel *`x0`*, controlled at a more stringent criterion of per-collection false positive of `<=0/31`, detects substantially less true-positives.  In other words, compared to panel *`x1`*, panel *`x0`* loses substantial power to detect true-positive due to a more stringent control of false-positive.
-   Panels *`x2`* and *`x3`*, controlled at less stringent criteria of per-collection false positive of `<=2/31` and `<=3/31`, respectively, have almost identical operating characteristics as panel *`x1`*.  In other words, compared to panel *`x1`*, panels *`x2`* and *`x3`* have no improvement on detection of true-positive, at the cost of higher false-positive.

```{r}
#| fig-height: 3
#| fig-width: 6
#| fig-align: left
#| fig-cap: 'Operating characteristics of `panellist` *`z`*'
#| label: fig-panellist
z |> 
  autoplot()
```

## Selected Collections


The `S3` method dispatch `as_flextable.panellist()` displays the collections selected in each `panel`.  The collections that contain more than one variants are highlighted using the [light-yellow]{style="background-color: lightyellow"} color.

```{r}
z |> 
  as_flextable()
```


# Appendix

## About {#sec-about}

R terminology may differ from that commonly used in mathematics and statistics. For explanations and references to the terms and abbreviations used in this vignette, please refer to @sec-terms.

### Environment {#sec-envir}

This vignette is created under `r R.version$version.string` with the aid of packages **`knitr`** [@knitr, v`r packageVersion('knitr')`], **`quarto`** [@quarto, v`r packageVersion('quarto')` with [Quarto](https://quarto.org/docs/get-started/) v`r quarto::quarto_version()`] and **`rmarkdown`** [@rmarkdown, v`r packageVersion('rmarkdown')`]. Documentation is supported by package **`roxygen2`** [@roxygen2, v`r packageVersion('roxygen2')`].

An Integrated Development Environment (IDE), e.g., [RStudio](https://posit.co/download/rstudio-desktop/) [@RStudio] or [Positron](https://positron.posit.co/download.html), is not required, but highly recommended.

### Dependency {#sec-depend}

Package **`minPanel`** `Imports` packages

-   **`cli`** [@cli, v`r packageVersion('cli')`], for attractive command line interfaces
-   **`consort`** [@consort, v`r packageVersion('consort')`], for Consolidated Standards of Reporting Trials (CONSORT) diagram
-   **`flextable`** [@flextable, v`r packageVersion('flextable')`], for flexible table
-   **`ggplot2`** [@ggplot2], **`patchwork`** [@patchwork], for visualization
-   `r '\U1f5dd'` **`matrixStats`** [@matrixStats, v`r packageVersion('matrixStats')`], key dependency, for matrix algebra
-   `r '\U1f5dd'` **`methods`** shipped with vanilla `r R.version.string`, key dependencies, for `S4` generic function and method dispatches
-   **`scales`** [@scales, v`r packageVersion('scales')`], for colour pallet


## `panel`

Function `panel()` (@sec-panel) returns an object of `S4` class `'panel'`, with `slot`s

-   `@m1`, a `logical` `matrix`, collections on rows and positive subjects on columns
-   `@m0`, a `logical` `matrix`, collections on rows and negative subjects on columns
-   `@id`, a `list` of `character` `vector`s.  Each `character` `vector` contains the variants in a collection. 
-   `@label`, an optional `character` scalar

The `S4` method dispatch of `show` to `signature = 'panel'` prints the vital information, e.g., the number of collections, and the number of positive and negative subjects.

### (Cumulative) True & False Positives {#sec-positive}

Function `true_positive()` finds the number of true positives per collection, i.e., the `rowSums` of the slot `@m1`.

Function `false_positive()` finds the number of false positives per collection, i.e., the `rowSums` of the slot `@m0`.

Functions `true_positive()` and `false_positive()` return 
an `integer` `vector`.
 
```{r}
# example
```


Function `cum_true_positive()` finds the number of true positives, by a sub-panel of the first $i$ collections.

Function `cum_false_positive()` finds the number of false positives, by a sub-panel of the first $i$ collections.

Function `cum_true_positive()` and `cum_false_positive()` return
a not-strictly increasing `integer` `vector`.

```{r}
# example
```


### Subset {#sec-subset}  

The `S3` method dispatch `subset.panel()` 🚧

#### by Panel-Wise False Positives

Panel *`x1`* in @sec-select-panel, controlled at `false_positive <= 1L` per collection, has a substantial panel-wise false positive rate

```{r}
x1
```

User may impose an additional criterion of panel-wise false positive of 5 or less out of 31 negative subjects.

```{r}
#| code-fold: true
#| code-summary: '**Example**: additional criteria of panel-wise false positive `<=5`'
x1a = m |>
  subset(subset = true_positive > 0L) |>
  subset(subset = false_positive <= 1L, append.label = TRUE) |>
  sort_by(y = ~ true_positive, decreasing = TRUE) |>
  subset(subset = diff(cum_true_positive) > 0L) |>
  subset(subset = cum_false_positive <= 5L, append.label = TRUE)
```

```{r}
x1a
```

@fig-panel-a shows ..

```{r}
#| code-fold: true
#| code-summary: '**Figure**: operating characteristics of *`x0`*, *`x1`* and *`x1a`*'
#| fig-height: 3
#| fig-width: 6
#| fig-align: left
#| label: fig-panel-a
#| fig-cap: 'Operating characteristics of *`x0`*, *`x1`* and *`x1a`*'
panellist(x0, x1, x1a) |>
  autoplot()
```


### Sort-By {#sec-sortby}

The `S3` method dispatch `sort_by.panel()` 🚧

### Append Label {#sec-appendlabel}

Function `append_label()` 🚧

### CONSORT Diagram {#sec-consort_panel}


Function `consort.panel()` creates a CONSORT diagram of the creation of a `panel`, using package **`consort`** [@consort].  

Function `consort.panel()` is a wrapper of the function `consort::consort_plot()`, which returns an object of class `'consort'`.

Examples using function `consort.panel()` are provided in @sec-select-panel.

## `panellist`

### CONSORT Diagram {#sec-consort_panellist}

Function `consort.panellist()` organizes the CONSORT diagrams of each `panel` member in a `panellist` in a collage, using packages **`consort`** [@consort] and **`patchwork`** [@patchwork].  

@fig-consort presents side-by-side the selection of collections in *`x0`*, *`x1`*, and *`x2`*.

```{r}
#| code-fold: true
#| code-summary: '**Figure**: CONSORT diagrams of *`x0`*, *`x1`*, and *`x2`*'
#| fig-height: 3.5
#| fig-width: 10
#| fig-align: left
#| fig-cap: 'CONSORT Diagram of `panellist`'
#| label: fig-consort
z[1:3] |> 
  consort.panellist(nrow = 1L)
```


# What We Don't Do

## `subset.panel()`

### by Multiple Criteria

too difficult to parse..

```{r}
#| eval: false
m |> 
  subset.panel(subset = (true_positive > 0L) & (false_positive <= 3L))
```


# Terms & Abbreviations {#sec-terms}


@tbl-Rterm presents a comprehensive glossary of R terms and abbreviations used in this vignette, which is folded to avoid overwhelming the reader with its excessive length.

R terminology and nomenclature could be *drastically different* from that of mathematics and statistics. Readers are *strongly advised* to read closely from the reference links in @tbl-Rterm, most of which point to webpages on [search.r-project.org](https://search.r-project.org), [cran.r-project.org](https://cran.r-project.org), [adv-r.hadley.nz](https://adv-r.hadley.nz) and [en.wikipedia.org](https://en.wikipedia.org).

<details>

<summary>@tbl-Rterm: Terms & Abbreviations in R</summary>

| Term / Abbreviation | Description |
|:-----------------------------------|:-----------------------------------|
| CRAN, R | The Comprehensive R Archive Network, <https://cran.r-project.org> |
| `Depends`, `Imports`, `Suggests`, `Enhances` | [*Writing R Extensions*, Section 1.1.3 *Package Dependencies*](https://cran.r-project.org/doc/manuals/r-devel/R-exts.html#Package-Dependencies) |
| [`|>`](https://search.r-project.org/R/refmans/base/html/pipeOp.html) | Forward pipe operator introduced since R version 4.1.0 |
| [`~`](https://search.r-project.org/R/refmans/base/html/tilde.html) | Tilde operator, to separate the left- and right-hand sides in a `formula` |
| `.` | Dot symbol, to denote the remaining variables contained in a [`data.frame`](https://search.r-project.org/R/refmans/base/html/data.frame.html). See [`formula`](https://search.r-project.org/R/refmans/stats/html/formula.html) and [`terms.formula`](https://search.r-project.org/R/refmans/stats/html/terms.formula.html) for details. |
| [`:`](https://search.r-project.org/R/refmans/base/html/Colon.html) | Colon operator for [`interaction`](https://search.r-project.org/R/refmans/base/html/interaction.html) of [`factor`](https://search.r-project.org/R/refmans/base/html/factor.html)s, see section *Details* of [`formula`](https://search.r-project.org/R/refmans/stats/html/formula.html) documentation |
| `/` | Forward slash operator for nested grouping structure, see section *Details* of [`formula`](https://search.r-project.org/R/refmans/stats/html/formula.html) documentation |
| [`::`, `:::`](https://search.r-project.org/R/refmans/base/html/ns-dblcolon.html) | Double and triple colon operators, explicitly-[namespace](https://search.r-project.org/R/refmans/base/html/ns-reflect.html)d function or object |
| [`args`](https://search.r-project.org/R/refmans/base/html/args.html) | Argument list of a [`function`](https://search.r-project.org/R/refmans/base/html/function.html) |
| [`attr`](https://search.r-project.org/R/refmans/base/html/attr.html), [`attributes`](https://search.r-project.org/R/refmans/base/html/attributes.html) | Attributes |
| [`body`](https://search.r-project.org/R/refmans/base/html/body.html) | Body of a [`function`](https://search.r-project.org/R/refmans/base/html/function.html) |
| [`data.frame`](https://search.r-project.org/R/refmans/base/html/data.frame.html) | Data frame |
| [`.Deprecated`](https://search.r-project.org/R/refmans/base/html/Deprecated.html), [`.Defunct`](https://search.r-project.org/R/refmans/base/html/Defunct.html) | R function retirement |
| [`deparse`, `deparse1`](https://search.r-project.org/R/refmans/base/html/deparse.html) | Expression deparsing, i.e., turn unevaluated expressions into `character` scalar or `vector` |
| [`diag`](https://search.r-project.org/R/refmans/base/html/diag.html) | Matrix diagonals |
| [`environment`](https://search.r-project.org/R/refmans/base/html/environment.html), [`as.environment`](https://search.r-project.org/R/refmans/base/html/as.environment.html) | (Coerce to an) environment |
| [`eval`](https://search.r-project.org/R/refmans/base/html/eval.html) | Evaluate an R `expression` |
| [`export`](https://cran.r-project.org/web/packages/roxygen2/vignettes/namespace.html) | Export tags in [NAMESPACE](https://bookdown.dongzhuoer.com/hadley/r-pkgs/namespace) |
| [`expression`](https://search.r-project.org/R/refmans/base/html/expression.html) | Unevaluated expressions |
| [`factor`](https://search.r-project.org/R/refmans/base/html/factor.html) | Categorical object, a.k.a., factor |
| [`file.size`](https://search.r-project.org/R/refmans/base/html/file.info.html) | File size in bytes |
| (one-sided) [`formula`](https://search.r-project.org/R/refmans/stats/html/formula.html) | Formula, or [one-sided formula](https://search.r-project.org/R/refmans/stats/html/asOneSidedFormula.html) |
| `foobar` | Placeholder; <https://en.wikipedia.org/wiki/Foobar> |
| [`formals`](https://search.r-project.org/R/refmans/base/html/formals.html) | Formal arguments of R function |
| [`function`](https://search.r-project.org/R/refmans/base/html/function.html) | R function definition |
| [`ggplot`](https://search.r-project.org/CRAN/refmans/ggplot2/html/ggplot.html) | Elegant data visualisations using the grammar-of-graphics, from package **`ggplot2`** [@ggplot2] |
| [`identical`](https://search.r-project.org/R/refmans/base/html/identical.html) | Exact equality in R |
| [`inherits`](https://search.r-project.org/R/refmans/base/html/class.html) | Class inheritance |
| [`Inf`, `is.finite`](https://search.r-project.org/R/refmans/base/html/is.finite.html) | Positive infinity $\infty$ |
| [`kable`](https://search.r-project.org/CRAN/refmans/knitr/html/kable.html) | Tables in LaTeX, HTML, markdown and reStructuredText, from package **`knitr`** [@knitr] |
| [`language`](https://search.r-project.org/R/refmans/base/html/is.language.html) | R language object |
| [`length`](https://search.r-project.org/R/refmans/base/html/length.html) | Length of an object |
| [`lengths`](https://search.r-project.org/R/refmans/base/html/lengths.html) | Lengths of `list` or `vector` elements |
| [`library`](https://search.r-project.org/R/refmans/base/html/library.html) | Loading/attaching packages |
| [`list`](https://search.r-project.org/R/refmans/base/html/list.html), [`listof`](https://search.r-project.org/R/refmans/stats/html/listof.html) | Lists of objects |
| [`list2env`](https://search.r-project.org/R/refmans/stats/html/list2env.html) | Create an `environment` containing all `list` components as objects |
| [`log`, `log1p`, `log2`, `log10`](https://search.r-project.org/R/refmans/base/html/Log.html) | $\log(x)$ and $\log(x+1)$ transformations |
| [`.Machine`](https://search.r-project.org/R/refmans/base/html/zMachine.html) | Numerical characteristics of the machine R is running on, e.g., 32-bit integers and IEC 60559 floating-point (double precision) arithmetic |
| [`match`](https://search.r-project.org/R/refmans/base/html/match.html) | Value matching |
| [`matrix`](https://search.r-project.org/R/refmans/base/html/matrix.html) | Matrices |
| [`merge`, `merge.data.frame`](https://search.r-project.org/R/refmans/base/html/merge.html) | Merge two `data.frame`s |
| [`mc.cores`](https://search.r-project.org/R/refmans/parallel/html/mclapply.html) | Number of central processing unit (CPU) cores to use for parallel computing |
| [`message`](https://search.r-project.org/R/refmans/base/html/message.html) | Diagnostic message printed in R console |
| [`missing`](https://search.r-project.org/R/refmans/base/html/missing.html) | R [`function`](https://search.r-project.org/R/refmans/base/html/function.html) not having a default value for a [`formals`](https://search.r-project.org/R/refmans/base/html/formals.html) argument |
| [storage `mode`](https://search.r-project.org/R/refmans/base/html/mode.html), [vector `mode`](https://search.r-project.org/R/refmans/base/html/vector.html) | Storage mode; vector mode |
| [`NaN`](https://search.r-project.org/R/refmans/base/html/is.finite.html) | Not-a-Number |
| [`na.fail`, `na.omit`, `na.exclude`, `na.pass`](https://search.r-project.org/R/refmans/stats/html/na.fail.html) | Handle missing values in R objects |
| [`names`](https://search.r-project.org/R/refmans/base/html/names.html) | Names of an R object |
| [`ncol`, `nrow`](https://search.r-project.org/R/refmans/base/html/nrow.html) | Number of rows/columns of an [`array`](https://search.r-project.org/R/refmans/base/html/array.html) |
| [`numeric`](https://search.r-project.org/R/refmans/base/html/numeric.html) | numeric, i.e., [`double`](https://search.r-project.org/R/refmans/base/html/double.html)-precision object |
| [`object.size`](https://search.r-project.org/R/refmans/utils/html/object.size.html) | Memory allocation |
| [`pi`](https://search.r-project.org/R/refmans/base/html/Constants.html) | Mathematical constant $\pi$ |
| `pmean`, `pmedian` | Point-wise mean and median, from package `groupedHyperframe` |
| [`pmax`, `pmin`](https://search.r-project.org/R/refmans/base/html/Extremes.html) | Point-wise maxima and minima |
| [`.Primitive`](https://search.r-project.org/R/refmans/base/html/Primitive.html) | *Primitive*, i.e., internally implemented, function |
| [`range`](https://search.r-project.org/R/refmans/base/html/range.html) | Range of values |
| [`return`](https://search.r-project.org/R/refmans/base/html/function.html) | R function return |
| `S3`, `generic`, [`UseMethod`](https://search.r-project.org/R/refmans/base/html/UseMethod.html), [`methods`](https://search.r-project.org/R/refmans/utils/html/methods.html), [`getS3method`](https://search.r-project.org/R/refmans/utils/html/getS3method.html) | `S3` object oriented system, <https://adv-r.hadley.nz/s3.html> |
| `S4`, [`isS4`](https://search.r-project.org/R/refmans/base/html/isS4.html), [`setClass`](https://search.r-project.org/R/refmans/methods/html/setClass.html), [`setGeneric`](https://search.r-project.org/R/refmans/methods/html/setGeneric.html), [`setMethod`](https://search.r-project.org/R/refmans/methods/html/setMethod.html), [`getMethod`](https://search.r-project.org/R/refmans/methods/html/getMethod.html) | `S4` object oriented system, <https://adv-r.hadley.nz/s4.html>, implemented in package **`methods`** shipped with `r R.version.string` |
| [`save`](https://search.r-project.org/R/refmans/base/html/save.html), [`saveRDS`](https://search.r-project.org/R/refmans/base/html/readRDS.html), `xz` | Save with [`xz`](https://en.wikipedia.org/wiki/XZ_Utils) compression |
| [`search`](https://search.r-project.org/R/refmans/base/html/search.html) | Search path |
| [`seq`, `seq.int`](https://search.r-project.org/R/refmans/base/html/search.html) | Sequence generation |
| [`signature`](https://search.r-project.org/R/refmans/methods/html/GenericFunctions.html) | Tools for managing `S4` generic functions |
| [`split`](https://search.r-project.org/R/refmans/base/html/split.html) | Divide into groups |
| [`stopifnot`](https://search.r-project.org/R/refmans/base/html/stopifnot.html) | To ensure the truth of R [`expression`](https://search.r-project.org/R/refmans/base/html/expression.html)s |
| [`symbol`, `name`](https://search.r-project.org/R/refmans/base/html/name.html) | R object name/symbol |
| [`t`, `t.default`](https://search.r-project.org/R/refmans/base/html/t.html) | Transpose |
| [`tempfile`](https://search.r-project.org/R/refmans/base/html/tempfile.html) | Create names for temporary files |
| [`typeof`](https://search.r-project.org/R/refmans/base/html/typeof.html) | R internal type or storage mode of any object |
| [`unclass`](https://search.r-project.org/R/refmans/base/html/class.html) | Remove `S3` `'class'` attribute |
| [`vector`](https://search.r-project.org/R/refmans/base/html/vector.html) | Vector |
| [`with`, `with.default`](https://search.r-project.org/R/refmans/base/html/with.html), [`with.hyperframe`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/with.hyperframe.html) | To `eval`uate an `expression` in a local `environment`, or in *each row* of a `hyperframe` |
| [`within`, `within.data.frame`](https://search.r-project.org/R/refmans/base/html/with.html) | To `eval`uate an `expression` in a local `environment` and makes modifications to the `environment` |

: Terms & Abbreviations in R {#tbl-Rterm}

</details>

{{< pagebreak >}}



# References {.appendix}

::: {#refs}
:::
